---
title: "Data Structures"
output: html_document
#output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Time estimate : 
#     Teaching : 45 mins
#     Exercise : 15 mins


##  Data Structures

#- "How can I load a package in R?"
#- "What is the 'tidyverse'?"
#- "How can I read data in R?"
#- "What are the basic data types in R?"

#objectives:

#- "To be aware of the different types of data."
#- "To begin exploring data frames, and understand how they are related to vectors, factors and lists."
# "To be able to ask questions from R about the type, class, and structure of an object."


# Note: it's nice to make a list in big writing on the board or taped to the wall
# listing all of these types and structures - leave it up for the rest of the workshop
# to remind people of the importance of these basics.
#  dataTypes <- c('double', 'complex', 'integer', 'character', 'logical')
# dataStructures <- c('data.frame', 'vector', 'factor', 'list', 'matrix')


```


One of R's most powerful features is its ability to deal with tabular data -
such as you may already have in a spreadsheet or a excel file. Let's start by
looking at a wetland dataset in your `data/` directory, called `WetPlots.xlsx`:

## The *readxl* package

To read the data into R, we are going to use our first package, called *readxl*. 
*readxl* is part of a suite of packages called the "tidyverse" which were designed
to work nicely together and to ease many common data operations.

The first time you use a package, you will need to install it (like installing
an app on your phone from the app store). Additionally, it is a good idea to
periodically check for updates to that package:

```{r, eval=FALSE}
install.packages("readxl")
```

Everytime we want to use that package, you must load it into your R session, by 
using the `library` function:

```{r}
library(readxl)
```

Now we can load this data into R via the following using the `read_excel` function, and assign it to an object called `wetland`:

```{r}

wetland <- read_excel("data/WetPlots.xlsx") 

```

The `read_excel` function is used for reading in tabular data from excel 
There are many other packages that do similar tasks (xlsx, xlsReadWrite), but the
 `readxl` version (`read_excel`) is a bit more user-friendly, and 
uses more sensible defaults.

The object that is created by `read_excel` is called a "tibble" - a rectangular
table-like object with rows and columns that is similar to a data frame.

We can begin exploring our dataset right away, first by looking at the whole thing:

```{r}
wetland
```

And pulling out individual columns by specifying them using the `$` operator:

```{r}
wetland$wetland_id
wetland$surveyors
wetland$newid
```

We can do other operations on the columns:

```{r}
## Say we wanted to increase the size of our id number by 10000
wetland$newid + 10000

paste("Site ", wetland$newid, "was surveyed by", wetland$surveyors)

```

But what about

```{r error=TRUE}
wetland$newid + wetland$surveyors
```

Understanding what happened here is key to successfully analyzing data in R.

## Data Types

If you guessed that the last command will return an error because site 1 with an id of `40076` plus
`"Don, Dustin, Neil"` is nonsense, you're right - and you already have some intuition for an
important concept in programming called *data types*. We can ask what type of
data something is:

```{r}
typeof(wetland$wetland_id)
```

There are 4 main types: 

- `double`/`numeric` (decimal numbers), 
- `integer` (counting numbers),
- `logical` (True/False),
- `character` (free text)

```{r}

typeof(3.14)
typeof(1L) # The L suffix forces the number to be an integer, since by default R uses double (decimal) numbers
typeof(TRUE)
typeof('banana')

```

No matter how
complicated our analyses become, all data in R is interpreted as one of these
basic data types. This strictness has some really important consequences.

The table that R loaded our wetland data into is something called a
*tibble*, and it is our first example of something called a *data
structure* - that is, a structure which R knows how to build out of the basic
data types.


```{r}
class(wetland)

```

## Vectors and Type Coercion

To better understand this behavior, let's meet another of the data structures:
the *vector*.

If we are creating vectors on our own, we will normally use the `c` (combine) 
function:

```{r}
my_vector <- c(1, 3, 5 ,7 ,9)
my_vector
```

A vector in R is essentially an ordered list of things, with the special
condition that *everything in the vector must be the same basic data type*.

```{r}
class(my_vector)
```

This command indicates the basic data type found in this vector - in this case `numeric`. 

We can use the logical operators that we learned earlier with vectors:

```{r}
my_vector > 4
```

Vectors can be any data type that we've already learned about. Let's make a 
character vector:

```{r}
my_other_vector <- c("Moose", "Bear", "Wolf", "Deer")
my_other_vector
```

```{r}
class(my_other_vector)
```

```{r}
my_other_vector == "Wolf"
```

If we similarly do

```{r}
class(wetland$newid)
```

we see that `class(wetland$newid)` is a vector, too - *the columns of data we load into R
data.frames are all vectors*, and that's the root of why R forces everything in
a column to be the same basic data type.

> ## Discussion 1
>
> Why is R so opinionated about what we put in our columns of data?
> How does this help us?
>

Given what we've learned so far, what do you think the following will produce?

```{r}
quiz_vector <- c(2,6,'3')
```

This is something called *type coercion*, and it is the source of many surprises
and the reason why we need to be aware of the basic data types and how R will
interpret them. When R encounters a mix of types (here numeric and character) to
be combined into a single vector, it will force them all to be the same
type. Consider:

```{r}
coercion_vector <- c('a', TRUE)
coercion_vector
another_coercion_vector <- c(0, TRUE)
another_coercion_vector
```

The coercion rules go: `logical` -> `integer` -> `double`/`numeric` -> `complex` ->
`character`, where -> can be read as *are transformed into*. You can try to
force coercion against this flow using the `as.` functions:

```{r}
character_vector_example <- c('0','2','4')
character_vector_example
character_coerced_to_numeric <- as.numeric(character_vector_example)
character_coerced_to_numeric
numeric_coerced_to_logical <- as.logical(character_coerced_to_numeric)
numeric_coerced_to_logical
```

As you can see, some surprising things can happen when R forces one basic data
type into another! Nitty-gritty of type coercion aside, the point is: if your
data doesn't look like what you thought it was going to look like, type coercion
may well be to blame; make sure everything is the same type in your vectors and
your columns of data.frames, or you will get nasty surprises!

But coercion can also be very useful! For example, in our `wetland` data
`slope_pc` is numeric, but we know that it also contains a "flat" which is not numeric. Therefore R will convert all the column into a `character` type. 

```{r}

wetland$slope_pc
as.numeric(wetland$slope_pc)

# note what happened to the "flat" site?

```



## Data Frames

We said that columns in data.frames were vectors:

```{r}
wetland$wetland_id
wetland$surveyors
```

## Factors

Another important data structure is called a *factor*. Factors usually look like
character data, but are typically used to represent categorical information that
have a defined set of values. For
example, let's make a vector of strings labelling the aspect in our study. This can only every be one of 8 catergories (N, NE, E, SE, S, SW, W, NW) and missing value (NA):

```{r}
aspect <- c('N', 'NE', 'E', 'SE', 'S', 'SW', 'W', 'NW')
aspect

```

We can turn a vector into a factor like so:

```{r}
aspect_class <- factor(aspect)
class(aspect_class)
aspect_class
```


## Factors - bonus details 

In modelling functions, it's important to know what the baseline levels are. This
is assumed to be the first factor, but by default factors are labelled in
alphabetical order. You can change this by specifying the levels:

```{r}
mydata <- c("case", "control", "control", "case")
factor_ordering_example <- factor(mydata, levels = c("control", "case"))
str(factor_ordering_example)
```

Now R has noticed that there are two possible categories in our data - but it
also did something surprising; instead of printing out the strings we gave it,
we got a bunch of numbers instead. R has replaced our human-readable categories
with numbered indices under the hood, this is necessary as many statistical
calculations utilise such numerical representations for categorical data.

In this case, we've explicitly told R that "control" should be represented by 1, and
"case" by 2. This designation can be very important for interpreting the
results of statistical models!

## Lists

Another data structure you'll want in your bag of tricks is the `list`. A list
is simpler in some ways than the other types, because you can put anything you
want in it:

```{r}
list_example <- list(1, "a", TRUE, 1+4i)
list_example
another_list <- list(title = "Numbers", numbers = 1:10, data = TRUE )
another_list
```

We can now understand something a bit surprising in our data.frame; what happens if we run:

```{r}
typeof(wetland)

```

We see that data.frames look like lists 'under the hood' - this is because a
data.frame is really a list of vectors and factors, as they have to be - in
order to hold those columns that are a mix of vectors and factors, the
data.frame needs something a bit more flexible than a vector to put all the
columns together into a familiar table.  In other words, a `data.frame` is a
special list in which all the vectors must have the same length.

In our `wetland` example, we have character and numeric variables 
we have seen already, each column of data.frame is a vector.

```{r}
wetland$newid
wetland[,2]
str(wetland[,2])
```

Each row is an *observation* of different variables, itself a data.frame, and
thus can be composed of elements of different types.

```{r}
wetland[1,]
typeof(wetland[1,])
str(wetland[1,])
```


<style>
div.green{ background-color:#FFFFE0; border-radius: 5px; padding: 20px;}
</style>
<div class = "green">

## Challenge 1

There are several subtly different ways to call variables, observations and
elements from data.frames:

- `wetland[1]`
- `wetland[[1]]`
- `wetland$fid`
- `wetland["fid"]`
- `wetland[1, 1]`
- `wetland[, 1]`
- `wetland[1, ]`

 Try out these examples and explain what is returned by each one.

 *Hint:* Use the function `typeof()` to examine what is returned in each case.

</div>

## Matrices

Last but not least is the matrix. We can declare a matrix full of zeros:

```{r}
matrix_example <- matrix(0, ncol=6, nrow=3)
matrix_example
```

And similar to other data structures, we can ask things about our matrix:

```{r}
class(matrix_example)
typeof(matrix_example)
str(matrix_example)
dim(matrix_example)
nrow(matrix_example)
ncol(matrix_example)
```

<style>
div.green{ background-color:#FFFFE0; border-radius: 5px; padding: 20px;}
</style>
<div class = "green">

## Challenge 2

What do you think will be the result of
 `length(matrix_example)`?
 Try it.
 Were you right? Why / why not?

## Challenge 3

 Make another matrix, this time containing the numbers 1:50,
 with 5 columns and 10 rows.
 Did the `matrix` function fill your matrix by column, or by
 row, as its default behaviour?
 See if you can figure out how to change this.
 (hint: read the documentation for `matrix`!)

## Challenge 4

Consider the R output of the matrix below:
```{r, echo=FALSE}
 matrix(c(4, 1, 9, 5, 10, 7), ncol = 2, byrow = TRUE)
```
 What was the correct command used to write this matrix? Examine
 each command and try to figure out the correct one before typing them.
 Think about what matrices the other commands will produce.

1. `matrix(c(4, 1, 9, 5, 10, 7), nrow = 3)`
2. `matrix(c(4, 9, 10, 1, 5, 7), ncol = 2, byrow = TRUE)`
3. `matrix(c(4, 9, 10, 1, 5, 7), nrow = 2)`
4. `matrix(c(4, 1, 9, 5, 10, 7), ncol = 2, byrow = TRUE)`

</div>


> ## Discussion 1
>
> By keeping everything in a column the same, we allow ourselves to make simple
> assumptions about our data; if you can interpret one entry in the column as a
> number, then you can interpret *all* of them as numbers, so we don't have to
> check every time. This consistency is what people mean when they talk about 
> *clean data*; in the long run, strict consistency goes a long way to making 
> our lives easier in R.
>
> ## Solution to Challenge 1
> ```{r, eval=TRUE, echo=TRUE}
> wetland[1]
> ```
> We can think of a data frame as a list of vectors. The single brace `[1]`
> returns the first slice of the list, as another list. In this case it is the
> first column of the data frame.
> ```{r, eval=TRUE, echo=TRUE}
>  wetland[[1]]
> ```
>  The double brace `[[1]]` returns the contents of the list item. In this case
> it is the contents of the first column, a _vector_ of type _factor_.
> ```{r, eval=TRUE, echo=TRUE}
>  wetland$fid
> ```
>  This example uses the `$` character to address items by name. _fid_ is the
> first column of the data frame, again a _vector_ of type _factor_.
> ```{r, eval=TRUE, echo=TRUE}
>  wetland["fid"]
> ```
> Here we are using a single brace `["fid"]` replacing the index number with
> the column name. Like example 1, the returned object is a _list_.
> ```{r, eval=TRUE, echo=TRUE}
>  wetland[1, 1]
> ```
>  This example uses a single brace, but this time we provide row and column
> coordinates. The returned object is the value in row 1, column 1. The object
> is an _integer_ but because it is part of a _vector_ of type _factor_, R
> displays the label "calico" associated with the integer value.
> ```{r, eval=TRUE, echo=TRUE}
>  wetland[, 1]
> ```
> Like the previous example we use single braces and provide row and column
> coordinates. The row coordinate is not specified, R interprets this missing
> value as all the elements in this _column_ _vector_.
> ```{r, eval=TRUE, echo=TRUE}
>  wetland[1, ]
> ```
> Again we use the single brace with row and column coordinates. The column
> coordinate is not specified. The return value is a _list_ containing all the
> values in the first row.
> 
> ## Solution to Challenge 2
> 
>  What do you think will be the result of
>  `length(matrix_example)`?
> 
> ```{r}
>  matrix_example <- matrix(0, ncol=6, nrow=3)
>  length(matrix_example)
> ```
> 
>  Because a matrix is a vector with added dimension attributes, `length`
>  gives you the total number of elements in the matrix.
> 
> 
> ## Solution to Challenge 3
> 
>  Make another matrix, this time containing the numbers 1:50,
> with 5 columns and 10 rows.
>  Did the `matrix` function fill your matrix by column, or by
>  row, as its default behaviour?
>  See if you can figure out how to change this.
>  (hint: read the documentation for `matrix`!)
> 
> ```{r, eval=FALSE}
>  x <- matrix(1:50, ncol=5, nrow=10)
>  x <- matrix(1:50, ncol=5, nrow=10, byrow = TRUE) # to fill by row
> ```
> 
> ## Solution to Challenge 4
> 
>  Consider the R output of the matrix below:
> ```{r, echo=FALSE}
>  matrix(c(4, 1, 9, 5, 10, 7), ncol = 2, byrow = TRUE)
> ```
>  What was the correct command used to write this matrix? Examine
>  each command and try to figure out the correct one before typing them.
>  Think about what matrices the other commands will produce.
> ```{r, eval=FALSE}
>  matrix(c(4, 1, 9, 5, 10, 7), ncol = 2, byrow = TRUE)
> ```
