---
title: "Dataframe Manipulation with dplyr"
output: html_document
#output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Time estimate : 
#     teaching: 40
#     exercises: 15

#questions:
#- "How can I manipulate dataframes without repeating myself?"
#objectives:
#- " To be able to use the six main dataframe manipulation 'verbs' with pipes in  `dplyr`."
#- " To understand how `group_by()` and `summarize()` can be combined to summarize datasets."
#- " Be able to analyze a subset of data using logical filtering."


```


Remembering that we can use the `readxl` package to read external data into R, for this lesson we are going to use the `wetland` data:

```{r}
library("readxl")
veg <- read_excel("data/WetPlots.xlsx", 3) 
veg
```


Manipulation of dataframes means many things to many researchers, we often
select certain observations (rows) or variables (columns), we often group the
data by a certain variable(s), or we even calculate summary statistics. We can
do these types of operations using the normal base R operations:

```{r}
mean(veg$gt_3m)
mean(veg$gt_3m, na = TRUE)

```

The question here is how might you take averages (or any other summary statistic) 
by some group you might be interested in.

## The `dplyr` package

Luckily, the [`dplyr`](https://cran.r-project.org/package=dplyr)
package provides a number of very useful functions for manipulating dataframes
in a way that will reduce the above repetition, reduce the probability of making
errors, and probably even save you some typing. As an added bonus, you might
even find the `dplyr` grammar easier to read.

Here we're going to cover 5 of the most commonly used functions as well as using
pipes (`%>%`) to combine them.

1. `select()`
2. `filter()`
3. `group_by()`
4. `summarize()`
5. `mutate()`

If you have have not installed this package earlier, please do so:

```{r,eval=FALSE}
install.packages('dplyr')
```

Now let's load the package:

```{r,message=FALSE}
library("dplyr")
```

## Using select()

If, for example, we wanted to move forward with only a few of the variables in
our dataframe we could use the `select()` function. This will keep only the
variables you select.

```{r}
veg_pc <- select(veg, plot.id, species.name, pc_cover)

```


## Using filter()

To select on some subset of rows:
```{r}
filter(veg, plot.id == 1)
```

Above we used 'normal' grammar, but the strengths of `dplyr` lie in combining 
several functions using pipes. This 
[description](https://twitter.com/WeAreRLadies/status/1172576445794803713) is very useful:

```{r, eval=FALSE}
I work up %>% 
  showered %>% 
  dressed %>% 
  had coffee %>% 
  came to an R workshop
```

Since the pipes grammar is unlike anything we've seen in R before, 
let's repeat what we've done above using pipes.

```{r}
veg_pc <- veg %>% select(plot.id, species.name, pc_cover)
```

To help you understand why we wrote that in that way, let's walk through it step
by step. First we summon the gapminder dataframe and pass it on, using the pipe
symbol `%>%`, to the next step, which is the `select()` function. In this case
we don't specify which data object we use in the `select()` function since in
gets that from the previous pipe. 

## Using filter() with pipes

If we now wanted to move forward with the above, but only with European
countries, we can combine `select` and `filter`

```{r}
veg_pc_p1 <- veg %>%
    filter(plot.id == 1) %>%
    select(species.name, pc_cover)
```


<style>
div.green{ background-color:#FFFFE0; border-radius: 5px; padding: 5px;}
</style>
<div class = "green">

## Challenge 1

 Write a single command (which can span multiple lines and includes pipes) that
 will produce a dataframe that has data for plot 2 and contains the column for `latin`, 
 `pc_cover`. How many rows does your dataframe
 have and why?

</div>
 
As with last time, first we pass the gapminder dataframe to the `filter()`
function, then we pass the filtered version of the gapminder dataframe to the
`select()` function. **Note:** The order of operations is very important in this
case. If we used 'select' first, filter would not be able to find the variable
plot.id since we would have removed it in the previous step.

## Using group_by() and summarize()

Now, we were supposed to be reducing the error prone repetitiveness of what can
be done with base R, but up to now we haven't done that since we would have to
repeat the above for each continent. Instead of `filter()`, which will only pass
observations that meet your criteria (in the above: `plot.id== 1`), we
can use `group_by()`, which will essentially use every unique criteria that you
could have used in filter.

```{r}
veg
veg %>% group_by(plot.id)

```
You will notice that the structure of the dataframe where we used `group_by()`
(`grouped_df`) is not the same as the original `veg` (`data.frame`). A
`grouped_df` can be thought of as a `list` where each item in the `list` is a
`data.frame` which contains only the rows that correspond to the a particular
value `plot.id` (at least in the example above).


## Using summarize()

The above was a bit on the uneventful side but `group_by()` is much more
exciting in conjunction with `summarize()`. This will allow us to create new
variable(s) by using functions that repeat for each of the plot-specific
data frames. That is to say, using the `group_by()` function, we split our
original dataframe into multiple pieces, then we can run functions
(e.g. `n()`, `mean()` or `sd()`) within `summarize()`.

```{r}
veg_sp_plot <- veg %>%
    group_by(fid) %>%
    summarize(count = n())
```


That allowed us to calculate the number of rows (species) per site, but it gets
even better.

<style>
div.green{ background-color:#FFFFE0; border-radius: 5px; padding: 5px;}
</style>
<div class = "green">

## Challenge 2

Calculate the average (mean) percent cover per plot. Which plot has the highest pc_cover? Hint calculations can only apply to numberic data types so you will need to convert the pc_cover 

</div>

                 
Another way to do this is to use the `dplyr` function `arrange()`, which
arranges the rows in a data frame according to the order of one or more
variables from the data frame.  It has similar syntax to other functions from
the `dplyr` package. You can use `desc()` inside `arrange()` to sort in descending order.

```{r}

veg$pc_cover <- as.numeric(veg$pc_cover)

plot_mean <- veg %>%
  group_by(plot.id) %>%
  summarize(mean_cover = mean(pc_cover, na.rm = TRUE))

plot_mean %>%
    arrange(mean_cover) %>%
    head(1)

plot_mean %>%
    arrange(desc(mean_cover)) %>%
    head(1)
```
 
The function `group_by()` allows us to group by multiple variables. Let's group by `plot.id` and `species.name`.


```{r}
cover_site_plot <- veg %>%
    group_by(fid, plot.id) %>%
    summarize(mean_pc_cover = mean(pc_cover, na.rm = TRUE))
```

That is already quite powerful, but it gets even better! You're not limited to defining 1 new variable in `summarize()`.

```{r}
cover_detail_pc <- veg %>%
    group_by(fid, plot.id) %>%
    summarize(mean_pc_cover = mean(pc_cover, na.rm = TRUE),
              sd_pc_cover = sd(pc_cover, na.rm = TRUE),
              mean_gt3m = mean(gt_3m, na.rm = TRUE),
              sd_gt3m = sd(gt_3m, na.rm = TRUE))
```

## count() and n()

A very common operation is to count the number of observations for each
group. The `dplyr` package comes with two related functions that help with this.

For instance, if we wanted to check the number of species included in the
dataset for one site, we can use the `count()` function. It takes the name
of one or more columns that contain the groups we are interested in, and we can
optionally sort the results in descending order by adding `sort=TRUE`:

```{r}
veg %>%
    filter(plot.id == 1) %>%
    count(species.name, sort = TRUE)
```

If we need to use the number of observations in calculations, the `n()` function
is useful. It will return the total number of observations in the current group rather than counting the number of observations in each group within a specific column. 

You can also chain together several summary operations; in this case calculating the `minimum`, `maximum`, `mean` and `se` of each continent's per-country life-expectancy:

```{r}
veg %>%
    group_by(fid) %>%
    summarize(
      mean_cover = mean(pc_cover, na.rm = TRUE),
      min_cover = min(pc_cover,na.rm = TRUE),
      max_cover = max(pc_cover, na.rm = TRUE))

```

## Using mutate()

We can also create new variables prior to (or even after) summarizing information using `mutate()`.

```{r}
library(stringr)

veg_groups <- veg %>%
  mutate(latin_family = word(latin, 1)) %>%
  group_by(fid, latin_family) %>%
  summarize(
    mean_cover = mean(pc_cover, na.rm = TRUE),
    min_cover = min(pc_cover,na.rm = TRUE),
    max_cover = max(pc_cover, na.rm = TRUE))

```

## Connect mutate with logical filtering: ifelse

When creating new variables, we can hook this with a logical condition. A simple combination of
`mutate()` and `ifelse()` facilitates filtering right where it is needed: in the moment of creating something new.
This easy-to-read statement is a fast and powerful way of discarding certain data (even though the overall dimension
of the data frame will not change) or for updating values depending on this given condition.

```{r}

veg_groups <- veg %>%
  mutate(latin_family = word(latin, 1)) %>%
  filter(latin_family == "carex") %>%
  group_by(fid ) %>%
  summarize(
    mean_cover = mean(pc_cover, na.rm = TRUE))

```

## Combining `dplyr` and `ggplot2`

First install and load ggplot2:

```{r,eval=FALSE}
install.packages('ggplot2')
```

```{r,message=FALSE}
library("ggplot2")
```

In the plotting lesson we looked at how to make a multi-panel figure by adding
a layer of facet panels using `ggplot2`. Here is the code we used (with some
extra comments):

```{r}
# plot average cover of all species per plot. 
# first we have to summarise the data 

cover_detail_pc <- veg %>%
    group_by(fid, plot.id) %>%
    summarize(mean_pc_cover = mean(pc_cover, na.rm = TRUE),
              sd_pc_cover = sd(pc_cover, na.rm = TRUE),
              mean_gt3m = mean(gt_3m, na.rm = TRUE),
              sd_gt3m = sd(gt_3m, na.rm = TRUE))

cover_detail_pc

ggplot(data = veg, aes(x = plot.id, y = pc_cover)) +
  geom_bar(stat = "identity") 


# Make the plot per site: 

ggplot(data = cover_detail_pc, aes(x = plot.id, y = mean_pc_cover)) +
  geom_bar(stat = "identity") + facet_wrap( ~ fid)


```


This code makes the right plot but it also provides a way to chain operations. Just as we used
`%>%` to pipe data along a chain of `dplyr` functions we can use it to pass data
to `ggplot()`. Because `%>%` replaces the first argument in a function we don't
need to specify the `data =` argument in the `ggplot()` function. By combining
`dplyr` and `ggplot2` functions we can alter this figure for only those continents in Europe.

```{r}

veg %>%
  group_by(fid, plot.id) %>%
  summarize(mean_pc_cover = mean(pc_cover, na.rm = TRUE),
            sd_pc_cover = sd(pc_cover, na.rm = TRUE),
            mean_gt3m = mean(gt_3m, na.rm = TRUE),
            sd_gt3m = sd(gt_3m, na.rm = TRUE)) %>%
  ggplot(aes(x = plot.id, y = mean_pc_cover)) +
  geom_bar(stat = "identity") + 
  facet_wrap( ~ fid)

```

Using `dplyr` functions also helps us do calculations on the fly. 

<style>
div.green{ background-color:#FFFFE0; border-radius: 5px; padding: 5px;}
</style>
<div class = "green">

## Advanced Challenge

Calculate the average pc_cover of 2 randomly selected species
for each fid. Then arrange the continent names in reverse order.
**Hint:** Use the `dplyr` functions `arrange()` and `sample_n()`, they have
similar syntax to other dplyr functions.

</div>


## Solution to Challenge 1
```{r}
veg_pc_p2 <- veg %>%
             filter(plot.id == 2) %>%
             select(latin, pc_cover)
```


## Solution to Challenge 2
```{r}
veg$pc_cover <- as.numeric(veg$pc_cover)

plot_mean <- veg %>%
  group_by(plot.id) %>%
  summarize(mean_cover = mean(pc_cover, na.rm = TRUE))

```
  
## Solution to Advanced Challenge
```{r}
 veg %>%
    group_by(fid) %>%
    sample_n(2) %>%
    summarize(mean_cover = mean(pc_cover)) %>%
    arrange(desc(mean_cover))
```


